Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community

### 2020-10-16 manuscript.rmd
1. Several instances of including co-authors suugestions 2. Moved main text Fig and table captions to the end of main text 3. Moved significant droughts figure to the main text 4. Moved goodness-of fit figures for obs vs. predict A and B fits to the main text

- Among the mechanisms that plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a species-specific leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of forest dynamics under current and future climate.

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycle [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity in mitigating mortality events. Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic vegetation models (DVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied. This mechanism assumes that water availability varies with soil depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

\
Trade-offs in plant 'traits' shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019].  Understanding how plant trade-offs lead to alternative life history strategies entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality).  Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011]. Acquisitive species have traits with high capacity to acquire, move, store and use water (higher hydraulic conductivity, low wood density) and achieve high growth, but build flimsy tissues (high specific leaf area) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more physiologically drought tolerant (greater embolism resistance, larger safety margins). Whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear. Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019]. Deep rooting may be expected to support an acquisitive strategy by enabling a steady water supply. However, the degree to which deep rooting is associated with an acquisitive strategy may depend on the specific dynamics of water availability by depth at a site, which is affected by precipitation regime and hydrology. For example, in pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (higher hydraulic conductivity) and opportunistic strategies (e.g. dry deciduousness) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].

\
Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. While correlations between aboveground hydraulic traits and mortality have been identified [@Anderegg:2016], numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]. However, plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.

\
To better understand how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, in this paper, we conduct a thorough assessment for a tropical forest of relationships between the water environment in the whole-soil column, above-ground hydraulic traits, estimates of rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration. Specifically, we test an overarching hypothesis that inter-specific variation in trees' rooting depths aligns species on an acquisitive-conservative economic spectrum, whereby acquisitive trees have rooting depths optimized for more reliable water availability by depth. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

\
```{r hypo, results = 'asis', echo = FALSE}
knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(hypo.table))))
```

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Descriptions of datasets used and details of some of the methods are described in the Supplementary Information.

\
In summary (Fig. \@ref(fig:schematic)), we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model. We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water dynamics best explained observed dynamics of species growth. This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves. For the latter, we collected new data on vulnerability curves for 21 species and scale-up to the community via scaling relationships with 'softer' traits, namely, Wood Density and Leaf Mass Area. We evaluated ERD against independent estimates based on existing data on plant and soil water isotopes. We then tested whether deeper ERD was associated with hydraulically efficient but vulnerable traits (Table 1 ), but not greater mortality rates.

```{r schematic, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-schematic}A schematic for the methods workflow.")}
knitr::include_graphics(file.path("figures/schematic.png"))
```

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Although 28% of canopy tree species are at least partly deciduous during the dry season, this amounts to the maximum percentage of deciduous individuals in the canopy this amounts to is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m x 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

## Model ELM-FATES for Soil Matric Potentials

@@

### Model parameterization
\
To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived using existing data on gravimetric water content vs. $\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential (200, mm H~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^ cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.

\
We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

\
We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we built with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).

@@

based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s. $W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.

Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensembles that minimised RMSE for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

## Model for Effective Rooting Depth (ERD)
\
Water is a key limiting resource for tree growth. Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD, in which a species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1 (see below). For the chosen model (see Model Evaluation below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrauligoical realisation $h$ is described as follows (see SI for equations for all the tested models):

@@

where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term. We estimated daily value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

@@

We estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $\rho$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

@@

As these fits explained large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(bci.AB$sp[!is.na(bci.AB$A)]))` species for which WSG and LMA were both available, retained `r length(unique(data.model.AB$sp))` for which parameters $A$ and $B$ were both positive, and fitted vulnerability curves to those (Eq. \#ref(eq:vc-exp)).

\
```{r symbols, results = 'asis', echo = FALSE}
knitr::kable(symbols.table, caption = 'The symbol, definition and units of key traits used.', booktabs = TRUE, escape = FALSE)
```

>>>>

- Among the mechanisms that plants employ to tolerate drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching critical safety thresholds. Here, we tested the role of deep-water access in enabling survival within a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD, as the average depth of water extraction), for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and to leaf hydraulic vulnerability curves. We validated ERD against existing tree and soil stable water isotope data. 
- Across species, deeper ERD was associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Species with deeper ERD spent less time below critical hydraulic thresholds indicating that trees compensate for water-stress related mortality risk through deeper water access.
- The role of deep-water access in alleviating mortality of hydraulically-vulnerable trees has important implications for our predictive understanding of forest dynamics under current and future climates.

@@

# Introduction
\
    Tropical forests play a disproportionately large role in the global carbon and energy cycles [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable among biomes given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity and hydraulic strategies in mitigating mortality events. Despite this, widespread drought-induced tree mortality has occurred across the tropics for specific functional groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic global vegetation models (DGVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell.2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
    Plants rely on a variety of structural and functional mechanisms to avoid or tolerate a drought, from adjustment of leaf area (deciduousness), change in leaf angle, reductions in stomatal conductance, upregulation of aquaporins, osmotic regulation to stem water storage, deep water access, increased root production, hydraulic redistribution and embolism resistance. Deep water access is arguably the most long-standing, yet under-studied mechanism. This mechanism assumes that plant available water varies with depth, and that the rooting depths of co-located trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018; @Brum.2019]. Consequently, trees in the same forest may differ in their experience during a drought, and thus in their growth and mortality responses [@Chitra-Tarak.2018]. A key bottleneck in community-wide testing of this mechanism has been a lack of data in both trees' rooting or water-sourcing depths and plant available soil water at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s more than 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though more than 90% of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in xylem water to variations in soil pore water with depth. However, such data are rare [@Evaristo:2016] and their interpretation is an active area of research given unresolved issues pertaining to soil heterogeneity and the correspondence of soil water extract in the laboratory to actual plant available water [@Adams.2020; @Bowers.2020; @Deurwaerder.2020]. DNA barcoding of roots [@Jones.2011] may be used to to estimate species-specific rooting depths or rooting profiles, although extent of sampling required may be cost-prohibitive and DNA barcode libraries for tropical forests are still under development. In general, community-scale data collection for rooting or water-sourcing depths is challenging for species-rich tropical forests.

\
    Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between aboveground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

\
    We used a locally parameterized and calibrated 1-D hydrological water balance model to obtain daily dynamics of soil water potential by depth in the whole-soil column and then applied a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water potential best explained observed dynamics of species-specific growth for large trees of `r length(erd.sp)` dominant species (Fig. \@ref(fig:schematic)). This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species-specific leaf vulnerability curves, by leveraging existing data for 8 species and developing trait-based proxies for the rest. We evaluated ERD against existing data on plant and soil water isotopes and tested coordination of ERD with hydraulic traits and mortality rates. Specifically, we tested an overarching hypothesis that inter-specific variation in trees' effective rooting depths (ERD) enables increasingly reliable water availability with depth and thus mitigation of hydraulic vulnerability. Our hypotheses are that (i) deep rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallow rooted species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deep rooted species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallow rooted species may have higher mortality rates as a result of higher variability water availability in shallow soil layers. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

Fig. \@ref(fig:schematic) here.
Table \@ref(tab:hypo) here.

# Materials and Methods
## Study site description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

\
In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm dbh size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48ha represent old growth forest that was never clear cut for agriculture. The plot (1000 m $\times$ 500 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot is situated in the Conrad catchment. Soils of the plot are homogeneous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot [@Baillie.2017]. The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere [@Croat.1978; @Leigh.1982; @Gentry.1990].

## Model ELM-FATES for soil matric potentials

@@

### Model parameterization
\
To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling (LHS) of this global parameter space. In the Conrad catchment, existing data on soil hydraulic conductivity for the depths of 12.5 and 60 cm showed a decline of conductivity by depth (@Kinner:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles) [Need to fix SI Figs. numbering to begin at 1]). For greater accuracy, instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\Psi_{soil}$ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted gravimetric water content to volumetric (VWC, $\theta$) using a bulk density value of of 0.8 g cm^-3^ measured at a point location (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

\begin{equation}
\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\Psi_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\Psi_e$, $\theta_{s}$ and $b$ taht best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

\
We ran ELM-FATES for the 5000-member parameter ensemble with hourly climate drivers from 1985-2018 [@Faybishenko.2020] initialized with the observed stand structure from the 50 ha plot [@Condit.2019a] in a single site mode. As our key interest here was in deriving soil water availability, we ran ELM-FATES in a lower-complexity configuration: static stand structure, and with a single plant functional type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of a dry-deciduous PFT did not significantly alter results (not shown). Holding the stand structure static rather than allowing it to emerge from physiologically-mediated competition between trees removes a set of internal model feedbacks and thus enables a cleaner experimental design where the only differences between model simulations are the direct physiological effects of perturbed model parameters.

### Model calibration
\
We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData.matteo)), (ii) stream discharge from the Conrad catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical TDR probe over 0-15 cm depth ($\theta_{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m (1299 samples in total covering all soil types and habitats; see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter, $\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ refer to plot-wide averages).

@@

based on monthly sums for both $ET$ and $q_{run}$ and daily averages for $\theta$s. $W_{liq,~\overline{z}}$ is the average of daily $W_{liq,~z}$ for $z$ $\in$ {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardised between 0 and 1 across the 5000 simulations.

Preliminary analyses showed that minimising the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit with soil moisture dynamics at 1 m depth, we first chose 300 parameter ensemble members that minimised RMSE between observations and simulations for VWC at the depth of 1m ($RMSE^{*}(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2), and among those 300 ensemble members we chose 100 members with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

## Model for effective rooting depth (ERD)
\
To estimate effective rooting depth, we tested alternate models for growth (Eqs. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many factors that could limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. In this study, we used maximum diurnal leaf conductance ($K_{leaf}$) approximated using rhizosphere soil water potential ($\Psi_{soil}$) as a proxy for the soil water limiting factor on growth. In our models $K_{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. For each species' growth dynamics across census intervals was regressed against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

For the chosen model (see Model validation and selection below), daily average growth for species $s$, $G_{s,~t~|~z}$, in the census interval $t$ for depth $z$ and for a given hydrological realisation $h$ is described as follows:

@@

where n~t~ be the total number of days in census interval *t*, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term. See SI for equations for all the alternate models tested.

As VPD may have a non-linear relationship with growth, with growth increasing for the lower range of VPD, and decreasing for the higher range of VPD due to stomatal closure (reference), we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

For each species, we estimated diurnal maximum value of $K_{leaf,~s,~i~|~z,~h}$ for a species $s$ by substituting $\Psi_{soil,~i~|~z,~h}$ in the leaf vulnerability curves defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$ as:

@@

By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit) [@Chonggang, we could have retained multiple depths with non-differentiable AICs], and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

@@

As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \#ref(eq:vc-exp). For all the 31 ERD species we estimated species $\Psi_{50,~leaf}$ based on their vulnerability curves (see Table \@ref(tab:symbols) for definition).

Table \@ref(tab:symbols) here.
